---
title: "Full Sample Analysis"
author: "Jennah Gosciak"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: false
---

```{r setup, set.seed(1690)}
```

```{r, include = F}
#options(warn=-1)
suppressPackageStartupMessages(library(tibble))
library(tidyverse)
library(knitr)
library(rgl)
library(rstan)
options(mc.cores = parallel::detectCores())
library(readr)
library(dplyr)
library(ggplot2)
library(tidyquant)
library(purrr)
library(rstanarm)
library(loo)
library(bayesplot)
library(haven)
library(sampleSelection)

set.seed(1690)
```

```{r}
## load data
df <- read_dta("../00_data/sample1.dta")
df <- df %>% 
  # create indicator for some college
  mutate(coll = if_else(str_sub(educus, 1, 1) == "8", 1, 0))

df_samp <- df
df_samp <- df_samp %>% 
  mutate(across(c("age", "age_fbirth"), ~ . - mean(., na.rm = T))) %>% 
  mutate(l_incwage = if_else(incwage <= 0, log(1), log(incwage)),
         l_wkswork1 = if_else(wkswork1 <= 0, log(1), log(wkswork1)))

df_samp %>% 
  group_by(samesex) %>% 
  summarize(n = n())
```

```{r}
# distribution of states
df_samp %>% 
  group_by(stateus) %>% 
  summarize(n = n()) %>% 
  arrange(desc(n))
```
```{r}
## distribution of education
df_samp %>% 
  group_by(coll) %>% 
  summarize(n = n())
```


```{r}
hist(df_samp$incwage, main = "Histogram of wage income")

hist(df_samp$l_incwage, main = "Histogram of log wage")
```

# Full Sample Analysis

## Running with `rstanarm` for initial testing

```{r, cache = TRUE}
# Running with rstanarm
post <-
  stan_glm(
    incwage ~ cnum_mt2 + age + age_fbirth + 
      f_boy + s_boy + r_black + hisp + r_oth,
    data = df_samp,
    family = gaussian(),
    prior = cauchy(),
    prior_intercept = cauchy(),
    seed = 12345
  )
post

post_log <-
  stan_glm(
    l_incwage ~ cnum_mt2 + age + age_fbirth + 
      f_boy + s_boy + r_black + hisp + r_oth,
    data = df_samp,
    family = gaussian(),
    prior = cauchy(),
    prior_intercept = cauchy(),
    seed = 12345
  )
post_log
```

## IV

* The likelihood function is based on 2.3 in the [sampleSelection vignette](https://cran.r-project.org/web/packages/sampleSelection/vignettes/treatReg.pdf)

```{r}
source(file.path("GLD_helpers.R"))
# set prior for rho with gld solver bounded
# allow rho to take on values between -1 and 1
a_s <- GLD_solver_bounded(bounds = -0.9:1, median = 0.3, IQR = 0.6)
r <- c(0.3, 0.6, a_s[1], a_s[2])
r

hist(qgld(runif(10000), r[1], r[2], r[3], r[4]), 
     main = "Prior distribution of rho")
```

```{r}
# display stan code
writeLines(readLines("iv_bin_fs.stan"))
```

```{r}
# subset the data
df_child <- df_samp %>% 
  filter(cnum_mt2 == 1)

df_nochild <- df_samp %>% 
  filter(cnum_mt2 == 0)

# reset covariates
cov <- c("age", "age_fbirth", "f_boy", "s_boy", "r_black", "hisp", "r_oth")

# set stan data
stan_data_iv <- list(N = nrow(df_samp),
                  N_child = nrow(df_child),
                  N_nochild = nrow(df_nochild),
                  K = 7,
                  X_child_s = df_child[, c("samesex", cov)],
                  X_nochild_s = df_nochild[, c("samesex", cov)],
                  X_child = df_child[, c(cov)],
                  X_nochild = df_nochild[, c(cov)],
                  y_child = df_child$l_incwage,
                  y_nochild = df_nochild$l_incwage,
                  prior_only = TRUE,
                  m = rep(-0.1, 5), 
                  scale = rep(0.3, 5), r = r)
```

```{r, cache = TRUE}
# call program without data for prior predictive checks
pre_iv <- stan("iv_bin.stan", data = stan_data_iv, seed = 1234)
```
```{r}
print(pre_iv, pars = 
        c("alpha0", "alpha1", "beta0", "beta1", "beta2", "sigma", "rho"))
```


```{r}
pairs(pre_iv, pars = c("alpha0", "beta0", "beta2", "sigma", "rho"))
```

The bivariate plots are primarily blobs and show no bivariate dependence relationships. Except for sigma and rho, the marginal distributions are approximately normal.

```{r}
stan_data_iv$prior_only <- FALSE
```
```{r, cache = TRUE}
post_iv <- stan("iv_bin_fs.stan", data = stan_data_iv, seed = 1234)
```
```{r}
print(post_iv, pars = c("alpha0", "alpha1", 
                        "beta0", "beta1", "beta2", "sigma", "rho"))
```

```{r}
pairs(post_iv, pars = c("alpha0", "beta2", "sigma", "rho"))
```

## Re-run with weeks worked outcome

```{r}
# set stan data
stan_data_iv <- list(N = nrow(df_samp),
                  N_child = nrow(df_child),
                  N_nochild = nrow(df_nochild),
                  K = 7,
                  X_child_s = df_child[, c("samesex", cov)],
                  X_nochild_s = df_nochild[, c("samesex", cov)],
                  X_child = df_child[, c(cov)],
                  X_nochild = df_nochild[, c(cov)],
                  y_child = df_child$wkswork1,
                  y_nochild = df_nochild$wkswork1,
                  prior_only = FALSE,
                  m = rep(-0.1, 5), 
                  scale = rep(1, 5), r = r)
```
```{r, cache = TRUE}
post_iv_wk <- stan("iv_bin_fs.stan", data = stan_data_iv, seed = 1234)
```
```{r}
print(post_iv_wk, pars = c("alpha0", "alpha1", 
                           "beta0", "beta1", "beta2", "sigma", "rho"))
```


## Re-run two-stage model for college subgroup

```{r}
# subset the data
df_samp2 <- df %>%
  filter(coll == 1) %>% 
  mutate(across(c("age", "age_fbirth"), ~ . - mean(., na.rm = T))) %>% 
  mutate(l_incwage = if_else(incwage <= 0, log(1), log(incwage)),
         l_wkswork1 = if_else(wkswork1 <= 0, log(1), log(wkswork1)))

df_child <- df_samp2 %>% 
  filter(cnum_mt2 == 1)

df_nochild <- df_samp2 %>% 
  filter(cnum_mt2 == 0)

# set stan data
stan_data_iv <- list(N = nrow(df_samp2),
                  N_child = nrow(df_child),
                  N_nochild = nrow(df_nochild),
                  K = 7,
                  X_child_s = df_child[, c("samesex", cov)],
                  X_nochild_s = df_nochild[, c("samesex", cov)],
                  X_child = df_child[, c(cov)],
                  X_nochild = df_nochild[, c(cov)],
                  y_child = df_child$l_incwage,
                  y_nochild = df_nochild$l_incwage,
                  prior_only = FALSE,
                  m = rep(-0.1, 5), 
                  scale = rep(0.3, 5), r = r)
```

```{r, cache = TRUE}
post_iv_coll <- stan("iv_bin_fs.stan", data = stan_data_iv, seed = 1234)
```
```{r}
print(post_iv_coll, pars = c("alpha0", "alpha1", 
                             "beta0", "beta1", "beta2", "sigma", "rho"))
```

## Conclusion

















